---
title: "LH Problem Set 1"
author: "Luke Haner"
date: "2025-09-29"
output:
  pdf_document: default
  word_document: default
---

```{r}

rm(list = ls())

knitr::opts_chunk$set(echo = TRUE)
setwd("C:/Users/lukeh/OneDrive/Desktop/Pols 602/Problem Set 1")
library(dplyr)
library(tidyr)
library(knitr)
library(ggplot2)
library(tinytex)
```


#Simulation Exercise 

#Randomly Samples N Observations from a population with some distribution of traits as N increases, and randomly assigns each observation to the treatment or control group with an equal probability 

```{r, echo=FALSE}

produce <- data.frame(
  Name = c('Strawberry', 'Blueberry', 'Orange', 'Banana'), 
  Color = c('Red', 'Blue', 'Orange', 'Yellow'),
  Frequency = c(0.30, 0.15, 0.45, 0.10) # proportions that equal 1 
)

repetition <- 100 #Number of repetitions 

set.seed(123)


#doing simulation once with fixed n at 1000
assignment_fixed <- replicate(repetition, { 
trait <- sample(produce$Name, size = 1000, replace = TRUE, prob = produce$Frequency) #creates vector of 1000 labels randomly chosen in line with population distribution 
Z <- rbinom(1000,1,0.5) # generates random draws from a binomial distribution with a coin flip probability to assign treatment and control groups  
data.frame(trait = trait, Z = Z) # combines two previous vectors (trait and Z) into a data frame 
}, simplify = FALSE)


assignment_many_times <- list() # container for sample sizes 
n_vals <-c(100, 300, 500, 700, 1000) #sample sizes as n increases 


for(n in n_vals){
  sims <- replicate(repetition,{
    trait <- sample(produce$Name, size=n, replace = TRUE, prob = produce$Frequency)
    Z <- rbinom(n, 1, 0.5) #assigns a coin flip probability from a normal distribution  
    data.frame(trait = trait, Z=Z)
  }, simplify = FALSE)
  
  assignment_many_times[[as.character(n)]] <- sims
  
}
```

#Calculate the proportion of traits for the entire sample, treatment, and control groups for each iteration.

```{r, echo=FALSE}

results <- data.frame(iter = integer(), group = character(), fruit_name = character(),prop = numeric()) #container for results 

#loop over each sample size in simulations
for (n in names(assignment_many_times)){ 

  sims <- assignment_many_times[[n]]

for (i in seq_along(sims)){
  df <- sims[[i]]
  n_obs <- nrow(df)
  
  #calculate proportions of each fruit in the full sample
  prop_all <- as.numeric(table(factor(df$trait, levels = produce$Name))) / n_obs   

  n_treat <- sum(df$Z ==1)
  n_ctrl <- n_obs - n_treat 
  #calculate proportion of each fruit in treatment group 
  prop_t <- as.numeric(table(factor(df$trait[df$Z == 1], levels = produce$Name))) / n_treat   
  #calculate proportion  of each fruit in the control group
  prop_c <- as.numeric(table(factor(df$trait[df$Z == 0], levels = produce$Name))) / n_ctrl   

  #Combine results into dataframe
  results <- dplyr::bind_rows(  
    results,
    data.frame(iter = i, n = as.numeric(n), group = 'All', fruit_name = produce$Name, prop = prop_all),
    data.frame(iter = i, n = as.numeric(n), group = 'Treat', fruit_name = produce$Name, prop = prop_t),
    data.frame(iter = i, n = as.numeric(n), group = 'Control', fruit_name = produce$Name, prop = prop_c)
    )
  }
}

#attach population proportions 
res_full <- results %>% 
  left_join(produce %>% rename(fruit_name = Name, pop_prop = Frequency),
            by = 'fruit_name') %>%
  mutate(imbalance = pop_prop - prop)
                     
#Table for visualization 
res_wide <- res_full %>% 
  select(-imbalance) %>% 
  pivot_wider(names_from = group, values_from = prop) %>%
  rename(population = pop_prop) %>%
  arrange(iter, fruit_name)

#Plot for Visualization
ggplot(results, aes(x = n, y = prop, color = group)) +  
  geom_point() + geom_line() +
  geom_hline(data = produce %>% rename(fruit_name = Name, pop_prop = Frequency), 
             aes(yintercept = pop_prop), linetype = 2) +
  facet_wrap(~ fruit_name) +
  labs(x = "Sample size (n)", y = "Proportion",
       title = "Treatment & Control proportions approach population as n increases",
       subtitle = "Dashed line = population proportion per fruit") +
  theme_minimal()


```
#Visualization Interpretation 
As the sample size increases it is observed that the proportions of all the fruit in the treatment and control groups move to be more and more in line with the population distribution (dashed line).
#Data Excercise
```{r}
voting_data <- read.csv("voting.csv")

```


#1.What is the treatment variable? Is it a discrete or continuous variable? What is the variable’s data type?
```{r}

typeof(voting_data$message)
```
The treatment variable is the message (Variable describing whether a voter received the social pressure message)

It is a discrete variable and has a character data type.


#2.Create a new treatment variable in your dataframe that is a binary version of the existing treatment variable. Your new variable should equal 1 if the observation was treated and 0 otherwise.
```{r}

message_binary <- ifelse(voting_data$message == 'yes', 1,0)

```

#3. Compute the average outcome of the treatment group and the average outcome for the control group. Interpret the results by writing 1-2 sentences about what these numbers mean substantively.
```{r}

average_treatment_outcome <- mean(voting_data$voted[message_binary == 1]) #takes average of treated outcomes 

average_control_outcome <- mean(voting_data$voted[message_binary ==0])# takes average of control outcomes 


print(average_treatment_outcome)
print(average_control_outcome)
```
The average outcome of the treatment group is 0.3779842, and the average outcome of the control group is 0.2966383. These numbers indicate the percentages of each group who went on to vote in an election. A larger amount of the treatment group went on to vote in an election compared to the control group, highlighting that the treatment led to an increased amount in voter turnout.



#4.Use brackets to subset the data frame and create two new data frames, one for the treatment group and one for the control group.
```{r}

treatment_group_df <- voting_data[message_binary == 1, ]

control_group_df <- voting_data[message_binary == 0, ]

head(treatment_group_df)
head(control_group_df)
```

#5.What is the average birth year for the treatment and control groups?
```{r}

treatment_average_birth_year <- mean(treatment_group_df$birth)
  
control_average_birth_year <- mean(control_group_df$birth)

print(treatment_average_birth_year)
print(control_average_birth_year)
```
The average birth year for the treatment group is 1956.147, and the average birth year for the control group is 1956.186.


#6.What is the estimated average causal effect for this experiment? Provide the calculated average effect and a substantive interpretation.
```{r}
average_causal_effect <- average_treatment_outcome - average_control_outcome
print(average_causal_effect)
```
The average causal effect is 0.08130991, meaning that the social pressure message caused an 8.1 percentage point increase in voter turnout compared to the control group who did not receive such a message.



#7.Suppose we wanted to claim that the estimated causal effect is an estimated effect for the entire U.S. population. What assumption would need to hold for us to make this claim?

The assumptions needed to generalize results to the entire U.S. population are that the control and treatment groups are comparable (in order to estimate average causal effects) and that the sample is representative of the overall U.S. population, best accomplished by random sampling individuals into the experiment and not excluding any group either intentionally or unintentionally.




